---
title: "Frequent A&E Attendance in Scotland: A Multilevel Analysis of Deprivation, Demographic, and Disease Burden Factors"
subtitle: "Objective 1b - Frequent Attenders"
author: "Fenanda Dwitha Kurniasari (S2744048)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    highlight: tango
    df_print: paged
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 10,
  fig.height = 6
)

# Global seed for reproducibility
set.seed(42)
```

# Packages

```{r packages}
library(tidyverse)
library(MASS)          # loaded for glm.nb() compatibility checks only; glmmTMB is used for all reported models
library(broom)
library(broom.mixed)
library(glmmTMB)
library(dplyr)
library(performance)   # icc(), check_collinearity()
library(DHARMa)
library(scales)
library(janitor)
library(readxl)
library(corrplot)
library(sf)
library(patchwork)
library(conflicted)

# MASS exports its own select() (used internally by stepAIC-style model
# selection), which silently masks dplyr::select() once MASS is loaded
# after tidyverse. Rather than rely on remembering to write dplyr::select()
# everywhere, we resolve the conflict once here so any accidental bare
# select()/filter()/rename() call fails loudly at the point of first use
# instead of throwing a confusing "unused arguments" error downstream.
conflicted::conflicts_prefer(dplyr::select, dplyr::filter, dplyr::rename, .quiet = TRUE)

theme_set(theme_minimal(base_size = 12))
eda_theme <- theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10, colour = "grey40", margin = margin(b = 8)),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(colour = "grey30", fill = NA, linewidth = 0.6),
    axis.title = element_text(size = 10.5)
  )

```

------------------------------------------------------------------------

# Data and Exploration

## Data Sources (2.1)

Three sources feed this analysis: PHS Multiple Attendances Demographics (the outcome, one row per stratum), PHS General Practice disease prevalence at Health Board level, and NRS mid-year 2025 population estimates used as the model offset.

```{r data-sources}
#PHS Multiple Attendances Demographics (rolling 12 months to April 2026)
#one row per stratum (hbt x age x sex x deprivation x department_type)
multiple_attendance <- read.csv("Data/multiple attendance_demographic.csv",
                                na.strings = c("", "NA")) %>% clean_names()

#PHS General Practice disease prevalence, Health Board level, 2025
disease_prevalence_raw <- read_excel("Data/diseaseprevalenceingeneralpractice_board_total.xlsx",
                                     sheet = "DP_Board_total") %>% clean_names()

# spatial data
hb_boundaries <- st_read("Data/SG_NHS_HealthBoards_2019/SG_NHS_HealthBoards_2019.shp")

# Area name -> Health Board code (hbt)
hb_name_to_code <- tibble::tribble(
  ~gp_practice_area, ~hbt,
  "NHS Ayrshire and Arran", "S08000015",
  "NHS Borders", "S08000016",
  "NHS Dumfries and Galloway","S08000017",
  "NHS Fife", "S08000029",
  "NHS Forth Valley",  "S08000019",
  "NHS Grampian","S08000020",
  "NHS Greater Glasgow and Clyde", "S08000031",
  "NHS Highland","S08000022",
  "NHS Lanarkshire","S08000032",
  "NHS Lothian", "S08000024",
  "NHS Orkney", "S08000025",
  "NHS Shetland", "S08000026",
  "NHS Tayside","S08000030",
  "NHS Western Isles","S08000028"
)

# ten candidate conditions (kept wide here, narrowed after the
# correlation/VIF check)
hbt_disease_prevalence <- disease_prevalence_raw %>%
  filter(area_type == "Board", age == "All", sex == "All") %>%
  left_join(hb_name_to_code, by = "gp_practice_area") %>%
  filter(!is.na(hbt)) %>%
  transmute(
    hbt, year,
    rate_copd  = rate_chronic_obstructive_pulmonary_disease_copd,
    rate_chd  = rate_coronary_heart_disease_chd,
    rate_heartfailure = rate_heart_failure,
    rate_diabetes  = rate_diabetes,
    rate_ckd   = rate_chronic_kidney_disease_ckd,
    rate_epilepsy  = rate_epilepsy,
    rate_dementia  = rate_dementia,
    rate_mentalhealth = rate_mental_health,
    rate_depression = rate_depression,
    rate_palliative = rate_palliative_care
  )

# NRS mid-year 2025 population, by Health Board x age x sex, the finest
# grain available (no deprivation or department-type split exists, see in in 3.1)
population_raw <- read_excel("Data/data-mid-year-population-estimates-2025.xlsx",
                             sheet = "Table 1", skip = 3) %>% clean_names() %>% mutate(year=2025L) 

age_col_map <- population_raw %>% dplyr::select(matches("^x[0-9]+(_and_over)?$")) %>% names()

population_long <- population_raw %>%
  dplyr::filter(area_type == "Health board", sex %in% c("Females", "Males")) %>%
  dplyr::select(year, area_name, area_code, sex, all_of(age_col_map)) %>%
  pivot_longer(cols = all_of(age_col_map), names_to = "age_col", values_to = "population") %>%
  mutate(
    single_age = if_else(age_col == "x90_and_over", 90L,
                          as.integer(str_remove(age_col, "^x"))),
    # collapse single-year ages into the same six bands used in the model
    age = case_when(
      single_age < 18             ~ "Under 18",
      between(single_age, 18, 24) ~ "18-24",
      between(single_age, 25, 39) ~ "25-39",
      between(single_age, 40, 64) ~ "40-64",
      between(single_age, 65, 74) ~ "65-74",
      single_age >= 75            ~ "75 plus"
    )
  )

hbt_age_sex_population <- population_long %>%
  group_by(hbt = area_code, age, sex, year) %>%
  summarise(population = sum(population), .groups = "drop") %>%
  mutate(age = factor(age, levels = c("Under 18", "18-24", "25-39",
                                       "40-64", "65-74", "75 plus")))
```

## Data Preprocessing (2.2)

### Deriving Frequent Attendance (2.2.1)

Frequent attendance is derived directly from Equations 2.1-2.2 in the report: `Total = n1+n2+n3+n4+n5plus`, `Frequent = n3+n4+n5plus`. The 3+ threshold follows Reid et al. (2025) and common practice in the literature, sitting below PHS's own clinical monitoring definition (Section 2.2.1 of the report explains why a three-month clustering threshold isn't supportable with this data).

### Data cleansing (2.2.2)

Missing age/sex/deprivation are checked before any exclusion, since the pattern of missingness (Figure 1) motivates treating this as MNAR rather than a simple data-entry gap.

```{r missing-data}
# Figure 1 - deprivation (17.1%) missing far more than age/sex (1.4% each);
# missing deprivation reflects no fixed Scottish postcode
# (homeless/tourists/short-term visitors), so this is MNAR not MCAR
missing_summary <- multiple_attendance %>%
  dplyr::select(age, sex, deprivation, department_type,
                one_attendance, two_attendances, three_attendances,
                four_attendances, five_plus_attendances) %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") %>%
  mutate(pct_missing = n_missing / nrow(multiple_attendance) * 100) %>%
  arrange(desc(pct_missing))

print(missing_summary)

ggplot(missing_summary, aes(x = reorder(variable, pct_missing), y = pct_missing)) +
  geom_col(fill = "firebrick", width = 0.65) +
  geom_text(aes(label = paste0(round(pct_missing, 1), "%")), hjust = -0.15, size = 3.4) +
  coord_flip(clip = "off") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Missing data by variable (raw multiple-attendance file)",
       x = NULL, y = "% missing") +
  eda_theme
ggsave("Figures/fig1_missing_data.png", width = 7.5, height = 4.5, dpi = 300)
```

We proceed under a stated MAR simplification and analyse complete cases (Section 2.2.2 of the report). Imputation was not attempted, since assigning a deprivation value to people without a fixed address would fabricate a profile for them.

```{r build-freq-data}
# complete-case exclusion + derive frequent/total (Eqns 2.1-2.2) + join
# disease prevalence and population
freq_data_prefilter <- multiple_attendance %>%
  filter(!is.na(age), !is.na(sex), !is.na(deprivation)) %>%
  mutate(
    age = factor(age, levels = c("Under 18", "18-24", "25-39",
                                  "40-64", "65-74", "75 plus")),
    sex = factor(sex),
    sex_pop = ifelse(sex == "Female", "Females", "Males"),
    deprivation = factor(deprivation, levels = 1:5,
                          labels = c("1 (most deprived)", "2", "3", "4",
                                     "5 (least deprived)")),
    department_type = factor(department_type),
    total_individuals = one_attendance + two_attendances + three_attendances +
                         four_attendances + five_plus_attendances,
    frequent = three_attendances + four_attendances + five_plus_attendances,
    prop_frequent = frequent / total_individuals
  ) %>%
  filter(total_individuals > 0) %>%
  left_join(hbt_disease_prevalence %>% filter(year == 2025), by = "hbt") %>%
  left_join(hbt_age_sex_population %>% filter(year == 2025) %>% rename(sex_pop = sex),
            by = c("hbt", "age", "sex_pop"))

# check, not silently drop, rows losing their population match at the join
n_dropped_join <- sum(is.na(freq_data_prefilter$population))
cat("Rows dropped due to unmatched population join:", n_dropped_join, "\n")

freq_data <- freq_data_prefilter %>% filter(!is.na(population), population > 0)
cat("Final analysis strata (n):", nrow(freq_data), "\n")
```

## Exploratory Data Analysis (2.3)

### Distribution of Frequent Attendance (2.3.1)

The mean/variance comparison here is the direct justification for negative binomial over Poisson, used formally again in Section 3.1.

```{r fig2-distribution}
dispersion_summary <- freq_data %>%
  summarise(n = n(), mean = mean(frequent), variance = var(frequent),
            sd = sd(frequent), var_mean_ratio = variance / mean)
print(dispersion_summary)

ggplot(freq_data, aes(x = frequent)) +
  geom_histogram(binwidth = 20, fill = "steelblue", colour = "white") +
  geom_vline(xintercept = dispersion_summary$mean, colour = "firebrick",
             linetype = "dashed", linewidth = 0.8) +
  annotate("label", x = dispersion_summary$mean, y = Inf,
           label = paste0("Mean = ", round(dispersion_summary$mean, 1)),
           colour = "firebrick", fill = "white", size = 3.4,
           vjust = 1.5, hjust = -0.1, label.size = 0.3) +
  annotate("label", x = max(freq_data$frequent, na.rm = TRUE) * 0.55, y = Inf,
           label = paste0(
             "n = ", dispersion_summary$n, "\n",
             "Mean = ", round(dispersion_summary$mean, 2), "\n",
             "SD = ", round(dispersion_summary$sd, 2), "\n",
             "Variance = ", round(dispersion_summary$variance, 2), "\n",
             "Var/Mean ratio = ", round(dispersion_summary$var_mean_ratio, 1)),
           hjust = 0, vjust = 1.1, size = 3.2, colour = "grey20",
           fill = "grey95", label.size = 0.3) +
  labs(title = "Distribution of frequent-attender counts per stratum",
       subtitle = paste0("n = ", dispersion_summary$n, " strata"),
       x = "Frequent attenders (count)", y = "Number of strata") +
  eda_theme
ggsave("Figures/fig2_frequent_distribution.png", width = 8, height = 5.5, dpi = 300)
```

### Demographic and Deprivation Patterns (2.3.2)

One combined figure only, deprivation x age x sex together, since that is what the report interprets: whether the deprivation gradient holds within every age/sex panel.

```{r fig3-demographics}
by_deprivation_age_sex <- freq_data %>%
  group_by(deprivation, age, sex) %>%
  summarise(frequent = sum(frequent), total = sum(total_individuals), .groups = "drop") %>%
  mutate(prop_frequent = frequent / total)

ggplot(by_deprivation_age_sex, aes(x = deprivation, y = prop_frequent, fill = sex)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.7) +
  geom_text(aes(label = scales::percent(prop_frequent, accuracy = 0.1)),
            position = position_dodge(width = 0.75),
            vjust = -0.4, size = 2.4, fontface = "bold", colour = "grey20") +
  scale_y_continuous(labels = scales::percent, expand = expansion(mult = c(0, 0.18))) +
  scale_fill_manual(values = c("Female" = "#F26D8B", "Male" = "#FDC08A")) +
  facet_wrap(~ age, nrow = 2) +
  labs(title = "Proportion of individuals with 3+ A&E attendances,\nby SIMD quintile, age group, and sex",
       x = "SIMD quintile", y = "% frequent attenders", fill = "Sex") +
  eda_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
        strip.text = element_text(face = "bold", size = 9),
        legend.position = "top")
ggsave("Figures/fig3_deprivation_age_sex.png", width = 10, height = 7, dpi = 300)
```

### Disease Prevalence and Frequent Attendance (2.3.3)

One correlation matrix does two jobs: shows the outcome association, and flags collinearity among the ten candidates (|r| > 0.7 is treated as a concern, Dormann et al. 2013).

```{r fig4-disease-correlation}
board_level_summary <- freq_data %>%
  group_by(hbt) %>%
  summarise(frequent = sum(frequent), total = sum(total_individuals), .groups = "drop") %>%
  mutate(prop_frequent = frequent / total) %>%
  left_join(hbt_disease_prevalence %>% filter(year == 2025), by = "hbt")

disease_cor_mat <- board_level_summary %>%
  dplyr::select(prop_frequent, starts_with("rate_")) %>%
  cor(use = "pairwise.complete.obs")

print(round(disease_cor_mat, 2))

png("Figures/fig4_disease_correlation.png", width = 950, height = 850, res = 150)
corrplot(disease_cor_mat, method = "color", type = "full",
         addCoef.col = "black", number.cex = 0.65,
         tl.col = "black", tl.srt = 45, tl.cex = 0.75,
         col = colorRampPalette(c("#4575B4", "white", "#D63C7E"))(200),
         title = "Pearson correlation: frequent attenders and\ncandidate disease-prevalence rates (Board level)",
         mar = c(0, 0, 3, 0))
dev.off()
```

```{r collinearity-check}
# collinearity check via VIF on a "kitchen sink" model with all ten rates 
# narrows the candidate list before the final model is fitted
obj1b_kitchensink <- glmmTMB(
  frequent ~ age + sex + deprivation + department_type +
    rate_copd + rate_chd + rate_heartfailure + rate_diabetes + rate_ckd +
    rate_epilepsy + rate_dementia + rate_mentalhealth + rate_depression +
    rate_palliative + offset(log(population)) + (1 | hbt),
  family = nbinom2, data = freq_data
)
check_collinearity(obj1b_kitchensink)

```

### Population Adjustment (2.3.4)

Rates, not raw counts, are what matter here; a large board like Greater Glasgow and Clyde would otherwise always look worse than NHS Orkney simply from population size.

```{r fig5-population-map}
hbt_total_population <- hbt_age_sex_population %>%
  filter(year == 2025) %>%
  group_by(hbt) %>%
  summarise(total_population = sum(population), .groups = "drop")

print(names(hb_boundaries))   # confirm code/name columns before joining

hb_pop_map <- hb_boundaries %>%
  left_join(hbt_total_population, by = c("HBCode" = "hbt"))

ggplot(hb_pop_map) +
  geom_sf(aes(fill = total_population), colour = "white", linewidth = 0.4) +
  geom_sf_text(aes(label = HBName), size = 2.6, colour = "grey15",
               fontface = "bold", check_overlap = TRUE) +
  scale_fill_gradientn(
    colours = c("#FFF6B7", "#FEE0A6", "#FDC08A", "#FA9070", "#F26D8B", "#D63C7E"),
    trans = "sqrt", labels = scales::comma,
    breaks = scales::pretty_breaks(n = 5), name = "Population\n(2025)"
  ) +
  labs(title = "Population Distribution Across\nNHS Scotland Health Boards",
       subtitle = "Mid-year population estimates, 2025",
       caption = "Source: National Records of Scotland") +
  theme_void(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, lineheight = 1.1),
    plot.subtitle = element_text(size = 10.5, colour = "grey40", hjust = 0.5, margin = margin(b = 10)),
    plot.caption = element_text(size = 8.5, colour = "grey50", hjust = 1),
    legend.title = element_text(face = "bold", size = 10),
    legend.text = element_text(size = 9),
    plot.margin = margin(20, 20, 15, 20)
  )
ggsave("Figures/fig5_population_map.png", width = 9, height = 8, dpi = 300)
```

------------------------------------------------------------------------

# Method

## Model Construction (3.1)

Negative binomial GLMM, `log(population)` offset, random intercept for `hbt`. Final disease block (heart failure, depression) selected from Figure 4 and the collinearity check above.

```{r model-final}
obj1b_model <- glmmTMB(
  frequent ~ age + sex + deprivation + department_type +
    rate_heartfailure + rate_depression +
    offset(log(population)) + (1 | hbt),
  family = nbinom2, data = freq_data
)
summary(obj1b_model)
```

The random intercept is justified via an LRT against a variance-boundary null (no `hbt` term); the fixed-effect set must match exactly, or `df_diff != 1`.

```{r random-effect-lrt}
obj1b_null <- glmmTMB(
  frequent ~ age + sex + deprivation + department_type +
    rate_heartfailure + rate_depression +
    offset(log(population)),
  family = nbinom2, data = freq_data
)

lr_stat_b <- as.numeric(2 * (logLik(obj1b_model) - logLik(obj1b_null)))
df_diff_b <- attr(logLik(obj1b_model), "df") - attr(logLik(obj1b_null), "df")
lr_p_b    <- pchisq(lr_stat_b, df = df_diff_b, lower.tail = FALSE) / 2
lr_stat_b; df_diff_b; lr_p_b

icc(obj1b_model)
```

## Model Selection (3.2)

Predictors were added in four blocks (age+sex, deprivation, department type, disease prevalence) rather than searched over, since a blind search would use the same data to choose and interpret the model. Each block is tested against the row above it via LRT and AIC (Table 1 of the report).

```{r model-selection-table1}
# M0: offset and (1|hbt) only
m0 <- glmmTMB(frequent ~ offset(log(population)) + (1 | hbt),
              family = nbinom2, data = freq_data)

# M1: + age, sex
m1 <- glmmTMB(frequent ~ age + sex + offset(log(population)) + (1 | hbt),
              family = nbinom2, data = freq_data)

# M2: + deprivation
m2 <- glmmTMB(frequent ~ age + sex + deprivation + offset(log(population)) + (1 | hbt),
              family = nbinom2, data = freq_data)

# M3: + department type
m3 <- glmmTMB(frequent ~ age + sex + deprivation + department_type +
                offset(log(population)) + (1 | hbt),
              family = nbinom2, data = freq_data)

# M4: + heart failure, depression rate (= obj1b_model, the final specification)
m4 <- obj1b_model

# each row tested against the row above via LRT (block-wise, not term-by-term)
lrt_block <- function(m_small, m_big) {
  lr <- as.numeric(2 * (logLik(m_big) - logLik(m_small)))
  df <- attr(logLik(m_big), "df") - attr(logLik(m_small), "df")
  p  <- pchisq(lr, df = df, lower.tail = FALSE)
  tibble(chisq = lr, df = df, p_value = p)
}

model_selection_table <- bind_rows(
  tibble(model = "M0", predictors = "Offset and (1|hbt) only",
         df = attr(logLik(m0), "df"), AIC = AIC(m0), chisq = NA, chisq_df = NA, p_value = NA),
  bind_cols(model = "M1", predictors = "M0 + age, sex",
            df = attr(logLik(m1), "df"), AIC = AIC(m1), lrt_block(m0, m1) %>% rename(chisq_df = df)),
  bind_cols(model = "M2", predictors = "M1 + deprivation",
            df = attr(logLik(m2), "df"), AIC = AIC(m2), lrt_block(m1, m2) %>% rename(chisq_df = df)),
  bind_cols(model = "M3", predictors = "M2 + department type",
            df = attr(logLik(m3), "df"), AIC = AIC(m3), lrt_block(m2, m3) %>% rename(chisq_df = df)),
  bind_cols(model = "M4", predictors = "M3 + heart failure, depression rate",
            df = attr(logLik(m4), "df"), AIC = AIC(m4), lrt_block(m3, m4) %>% rename(chisq_df = df))
)

model_selection_table %>%
  knitr::kable(digits = 2, caption = "Model comparison for selecting final predictors (Table 1).")

```

M4 is taken forward as the final specification. Diagnostics follow in Section A.1.

## Sensitivity Analysis for Final Model (3.3)

### Model Assumption and Diagnostic (Appendix A.1)

Simulation-based residual diagnostics from `DHARMa`, run on M4, avoid applying linear-model residual checks directly to count data.

```{r dharma-diagnostics}
obj1b_sim <- simulateResiduals(obj1b_model, n = 1000)

png("Figures/diag_dharma_overall.png", width = 10, height = 5, units = "in", res = 300)
plot(obj1b_sim)
dev.off()

png("Figures/diag_residuals.png", width = 8, height = 5, units = "in", res = 300)
plotResiduals(obj1b_sim)
dev.off()

png("Figures/diag_outliers.png", width = 6, height = 5, units = "in", res = 300)
testOutliers(obj1b_sim, type = "bootstrap")   # bootstrapped, preferred at this sample size
dev.off()

png("Figures/diag_dispersion.png", width = 6, height = 5, units = "in", res = 300)
testDispersion(obj1b_sim)
dev.off()

png("Figures/diag_zeroinflation.png", width = 6, height = 5, units = "in", res = 300)
testZeroInflation(obj1b_sim)   # this is the one that flags a genuine gap
dev.off()
```

KS test and dispersion test both pass; the zero-inflation test does not (M4 under-predicts zero-count strata). This motivates the ZINB check below, reported as a limitation, not a change to the primary model.

### Poisson vs NB vs ZINB comparison (Figures 6-7)

```{r sensitivity-fit-comparators}
pois_fit <- glmmTMB(frequent ~ age + sex + deprivation + department_type +
                      rate_heartfailure + rate_depression +
                      offset(log(population)) + (1 | hbt),
                    family = poisson, data = freq_data)

obj1b_zinb_dept <- glmmTMB(
  frequent ~ age + sex + deprivation + department_type +
    rate_heartfailure + rate_depression +
    offset(log(population)) + (1 | hbt),
  ziformula = ~ department_type,
  family = nbinom2, data = freq_data
)
```

```{r fig6-aic-zero-comparison}
# Panel 1 - AIC comparison
aic_df <- data.frame(
  model = c("Poisson", "Negative Binomial (M4)", "ZINB (department type)"),
  AIC = c(AIC(pois_fit), AIC(obj1b_model), AIC(obj1b_zinb_dept))
)
aic_df$model <- factor(aic_df$model, levels = aic_df$model)

p1 <- ggplot(aic_df, aes(x = model, y = AIC, fill = model)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = round(AIC, 0)), vjust = -0.5) +
  labs(title = "Model fit comparison (AIC)", x = NULL, y = "AIC") +
  theme_minimal()

# Panel 2 - observed vs model-expected zero-count strata
obs_zeros <- sum(freq_data$frequent == 0)

sim_pois <- simulateResiduals(pois_fit, n = 1000)
sim_nb   <- simulateResiduals(obj1b_model, n = 1000)
sim_zinb <- simulateResiduals(obj1b_zinb_dept, n = 1000)

exp_zeros <- function(sim) mean(colSums(sim$simulatedResponse == 0))

zero_df <- data.frame(
  model = c("Observed", "Poisson", "Negative Binomial (M4)", "ZINB (department type)"),
  zeros = c(obs_zeros, exp_zeros(sim_pois), exp_zeros(sim_nb), exp_zeros(sim_zinb))
)
zero_df$model <- factor(zero_df$model, levels = zero_df$model)

p2 <- ggplot(zero_df, aes(x = model, y = zeros, fill = model == "Observed")) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = round(zeros, 0)), vjust = -0.5) +
  scale_fill_manual(values = c("TRUE" = "firebrick", "FALSE" = "steelblue")) +
  labs(title = "Observed vs model-expected zero-count strata", x = NULL, y = "Number of strata") +
  theme_minimal()

p1 + p2   # Figure 6
ggsave("Figures/fig6_sensitivity_comparison.png", p1 + p2, width = 10, height = 5, dpi = 300)
```

```{r fig7-coefficient-comparison}
extract_coefs <- function(model, label) {
  tidy(model, effects = "fixed", conf.int = TRUE) %>%
    filter(component == "cond", term != "(Intercept)") %>%
    mutate(model = label)
}

coef_df <- bind_rows(
  extract_coefs(pois_fit,        "Poisson"),
  extract_coefs(obj1b_model,     "Negative Binomial (M4)"),
  extract_coefs(obj1b_zinb_dept, "ZINB (department type)")
)
coef_df$model <- factor(coef_df$model,
                        levels = c("Poisson", "Negative Binomial (M4)", "ZINB (department type)"))

ggplot(coef_df, aes(x = estimate, y = term, colour = model)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high),
                  position = position_dodge(width = 0.6), size = 0.4) +
  labs(title = "Fixed-effect coefficients across model specifications",
       x = expression(beta~"(log scale)"), y = NULL, colour = "Model") +
  theme_minimal() +
  theme(legend.position = "bottom")
ggsave("Figures/fig7_coef_comparison.png", width = 9, height = 7, dpi = 300)
```

M4 and the ZINB extension agree on almost every predictor. heart failure prevalence is the only coefficient that shifts meaningfully (Section 3.3 of the report).

### Zero-inflated Negative Binomial Model (Appendix A.2)

```{r appendix-a2-zinb-extension}
# zero-inflation conditioned on department type only - the variable most
# plausibly associated with structural zeros (Type 3 sites)
zinb_ratio_dept <- exp_zeros(sim_zinb) / obs_zeros

# also tested: zero-inflation ~ 1 (intercept-only) and ~ deprivation +
# department_type, kept here as a documented negative result
obj1b_zinb <- glmmTMB(
  frequent ~ age + sex + deprivation + department_type +
    rate_heartfailure + rate_depression +
    offset(log(population)) + (1 | hbt),
  ziformula = ~ 1,
  family = nbinom2, data = freq_data
)

# adding deprivation reintroduced a separation problem (huge SEs on
# deprivation2/4 in the zi component) - NOT carried forward, kept only
# as evidence for why the department-type-only extension was preferred
obj1b_zinb_dep_dept <- glmmTMB(
  frequent ~ age + sex + deprivation + department_type +
    rate_heartfailure + rate_depression +
    offset(log(population)) + (1 | hbt),
  ziformula = ~ deprivation + department_type,
  family = nbinom2, data = freq_data
)
summary(obj1b_zinb_dep_dept)   # inspect the inflated SEs on deprivation terms

AIC(obj1b_model, obj1b_zinb_dept)   # lower AIC = preferred, penalised for the extra zi parameter

zinb_comparison <- bind_rows(
  tidy(obj1b_model, effects = "fixed", exponentiate = TRUE, conf.int = TRUE) %>%
    mutate(model = "NB (main)"),
  tidy(obj1b_zinb, effects = "fixed", component = "cond",
       exponentiate = TRUE, conf.int = TRUE) %>%
    mutate(model = "ZINB (conditional)")
) %>%
  dplyr::filter(term != "(Intercept)") %>%
  dplyr::select(model, term, estimate, conf.low, conf.high, p.value)

print(zinb_comparison, n = Inf)


ggplot(zinb_comparison, aes(x = term, y = estimate, ymin = conf.low, ymax = conf.high, colour = model)) +
  geom_pointrange(position = position_dodge(width = 0.5)) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  coord_flip() +
  scale_colour_manual(values = c("NB (main)" = "steelblue", "ZINB (conditional)" = "firebrick")) +
  labs(title = "Sensitivity check: NB vs. zero-inflated NB",
       subtitle = "Coefficients pointing the same direction/magnitude support the NB model's robustness",
       x = NULL, y = "Rate ratio (95% CI)", colour = "Model") +
  eda_theme
ggsave("Figures/sensitivity3_zinb_comparison_plot.png", width = 9, height = 7, dpi = 300)

# flag any coefficient that meaningfully diverges (direction flip or
# non-overlapping CIs) - these need explicit discussion, not silent burial
zinb_flag <- zinb_comparison %>%
  dplyr::select(model, term, estimate) %>%
  pivot_wider(names_from = model, values_from = estimate) %>%
  mutate(
    direction_flip = sign(log(`NB (main)`)) != sign(log(`ZINB (conditional)`)),
    pct_diff = abs(`NB (main)` - `ZINB (conditional)`) / `NB (main)` * 100
  ) %>%
  arrange(desc(pct_diff))

print(zinb_flag, n = Inf)

```

The ratio of observed to expected zero-count strata falls from 1.94 (M4) to about 1.4 once department type enters the zero-inflation component (Section A.2 of the report), confirming department type explains most, not all, of the excess zeros. M4 is retained as the primary specification throughout.

### Robustness checks: disease block and attendance threshold

```{r sensitivity-disease-block}
# Check 1 - does the disease-prevalence block add anything on its own?
obj1b_no_disease <- glmmTMB(
  frequent ~ age + sex + deprivation + department_type +
    offset(log(population)) + (1 | hbt),
  family = nbinom2, data = freq_data
)

anova(obj1b_no_disease, obj1b_model)
AIC(obj1b_no_disease, obj1b_model)
```

```{r sensitivity-threshold}
# this one I didnt include in the report but worthwhile to check 
# Check 2, does the 3+ threshold change the story vs 4+ / 5+?
refit_freq_model <- function(threshold_data, label) {
  m <- glmmTMB(
    frequent ~ age + sex + deprivation + department_type +
      rate_heartfailure + rate_depression +
      offset(log(population)) + (1 | hbt),
    family = nbinom2, data = threshold_data
  )
  tidy(m, effects = "fixed", exponentiate = TRUE, conf.int = TRUE) %>% mutate(threshold = label)
}

freq_data_thresholds <- freq_data %>%
  mutate(frequent_3plus = three_attendances + four_attendances + five_plus_attendances,
         frequent_4plus = four_attendances + five_plus_attendances,
         frequent_5plus = five_plus_attendances)

sensitivity_threshold <- bind_rows(
  freq_data_thresholds %>% mutate(frequent = frequent_3plus) %>% refit_freq_model("3+ (main)"),
  freq_data_thresholds %>% mutate(frequent = frequent_4plus) %>% refit_freq_model("4+"),
  freq_data_thresholds %>% mutate(frequent = frequent_5plus) %>% refit_freq_model("5+")
)


ggplot(sensitivity_threshold %>% filter(term != "(Intercept)"),
       aes(x = term, y = estimate, ymin = conf.low, ymax = conf.high, colour = threshold)) +
  geom_pointrange(position = position_dodge(width = 0.5)) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  coord_flip() +
  labs(title = "Sensitivity of rate ratios to the frequent-attender threshold",
       x = NULL, y = "Rate ratio (95% CI)", colour = "Threshold") +
  eda_theme
ggsave("Figures/sensitivity2_threshold_plot.png", width = 9, height = 7, dpi = 300)
```

------------------------------------------------------------------------

# Result and Discussion

## Table 2: Fitted Coefficients and Rate Ratios

```{r table2-results}
table2_results <- tidy(obj1b_model, effects = "fixed", component = "cond",
                        exponentiate = FALSE, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(rate_ratio = exp(estimate),
         ci_low = exp(conf.low), ci_high = exp(conf.high)) %>%
  dplyr::select(term, estimate, rate_ratio, ci_low, ci_high, p.value)

table2_results %>%
  knitr::kable(digits = 3, caption = "Table 2. Fitted coefficients and rate ratios from the final negative binomial GLMM.")

```

Sections 4.1-4.4 of the report walk through these rate ratios block by block (age, sex, deprivation, department type, disease prevalence); no separate code is needed beyond the table above.

## Health Board Random Effect (4.5)

### Caterpillar plot (Figure 9)

```{r fig9-caterpillar}
# tidy() with ran_vals pulls the conditional variance needed for CIs,
# unlike ranef() alone
board_effects_ci <- tidy(obj1b_model, effects = "ran_vals", conf.int = TRUE) %>%
  filter(group == "hbt") %>%
  rename(hbt = level, log_effect = estimate) %>%
  mutate(
    rate_ratio = exp(log_effect),
    ci_low     = exp(conf.low),
    ci_high    = exp(conf.high),
    # significant if the 95% CI excludes 1 (genuinely differs from expected)
    significant = if_else(ci_low > 1 | ci_high < 1, "Significant", "Not significant")
  ) %>%
  arrange(rate_ratio) %>%
  mutate(hbt_rank = row_number())


hb_lookup <- hb_boundaries %>% st_drop_geometry() %>% dplyr::select(HBCode, HBName)

board_effects_ci <- board_effects_ci %>%
  left_join(hb_lookup, by = c("hbt" = "HBCode")) %>%
  mutate(hbt = factor(HBName, levels = HBName[order(rate_ratio)]))

ggplot(board_effects_ci, aes(x = hbt, y = rate_ratio, colour = significant)) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey40") +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high), width = 0, linewidth = 0.4, alpha = 0.7) +
  geom_point(size = 1.6) +
  scale_colour_manual(values = c("Significant" = "firebrick", "Not significant" = "steelblue")) +
  labs(
    title = "Adjusted Frequent-Attender Rate Ratios Across Health Boards",
    subtitle = "Ranked by rate ratio vs. model expectation, with 95% CI",
    x = "Health Board", y = "Rate ratio vs. expected", colour = NULL,
    caption = "After adjusting for age, sex, deprivation, department type, and disease prevalence"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 13, hjust = 0.5),
    plot.subtitle = element_text(size = 10, colour = "grey40", hjust = 0.5),
    plot.caption = element_text(size = 8, colour = "grey50"),
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8)
  )
ggsave("Figures/fig9_board_effects_caterpillar.png", width = 9, height = 6, dpi = 300)
```

Only Tayside, Forth Valley, and Dumfries and Galloway have CIs that exclude 1; the remaining eleven boards sit within the range chance would produce.

### Map (Figure 8)

```{r fig8-board-map}
board_effects <- ranef(obj1b_model)$cond$hbt %>%
  rownames_to_column("hbt") %>%
  rename(log_effect = `(Intercept)`) %>%
  mutate(rate_ratio_vs_expected = exp(log_effect),
         direction = if_else(log_effect > 0, "Higher than expected", "Lower than expected")) %>%
  arrange(desc(rate_ratio_vs_expected))


hb_map_data <- hb_boundaries %>%
  left_join(board_effects, by = c("HBCode" = "hbt")) %>%
  mutate(label_text = paste0(HBName, "\n", sprintf("%.2f", rate_ratio_vs_expected)))

ggplot(hb_map_data) +
  geom_sf(aes(fill = rate_ratio_vs_expected), colour = "white", linewidth = 0.3) +
  geom_sf_text(aes(label = label_text), size = 2.4, colour = "grey15",
               fontface = "bold", lineheight = 0.85, check_overlap = TRUE) +
  scale_fill_gradient2(
    low = "steelblue", mid = "white", high = "firebrick", midpoint = 1,
    labels = scales::number_format(accuracy = 0.01),
    name = "Rate ratio\nvs. expected"
  ) +
  labs(
    title = "Frequent-Attender Rate by Health Board: Observed vs. Model-Expected",
    subtitle = "Red = higher than expected; blue = lower than expected, after adjusting for\nage, sex, deprivation, department type, and disease prevalence",
    caption = "Values are rate ratios (1 = as expected). Boundaries: NHS Scotland Health Boards, 2019."
  ) +
  theme_void(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5, margin = margin(b = 4)),
    plot.subtitle = element_text(size = 9.5, colour = "grey40", hjust = 0.5, margin = margin(b = 10)),
    plot.caption = element_text(size = 8, colour = "grey50", hjust = 1, margin = margin(t = 10)),
    legend.position = "right",
    legend.title = element_text(face = "bold", size = 10),
    legend.text = element_text(size = 9),
    plot.margin = margin(15, 15, 15, 15)
  )
ggsave("Figures/fig8_board_effects_map.png", width = 9, height = 8, dpi = 300)
```

------------------------------------------------------------------------

# Appendix A.3: Age-by-Sex Interaction Model

The main text reports the additive model. This tests `age * sex` in place of the additive terms, as a robustness check reported in Appendix A.3 and revisited in the Limitations section.

```{r appendix-a3-interaction}
obj1b_model_int <- glmmTMB(
  frequent ~ age * sex + deprivation + department_type +
    rate_heartfailure + rate_depression +
    offset(log(population)) + (1 | hbt),
  family = nbinom2, data = freq_data
)

anova(obj1b_model, obj1b_model_int)
summary(obj1b_model_int)

interaction_table <- tidy(obj1b_model_int, effects = "fixed", component = "cond",
                           conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(rate_ratio = exp(estimate), ci_low = exp(conf.low), ci_high = exp(conf.high))

interaction_table %>%
  knitr::kable(digits = 3, caption = "Table 3. Age-by-sex interaction model, conditional model coefficients.")

```

Only the 18-24 x Male interaction is significant; the other four age bands show no departure from the additive pattern. The additive model (M4) is retained as the primary specification, and the young-men result is flagged for further work rather than folded into the general age effect (Section 6 of the report).

------------------------------------------------------------------------

*End of analysis. All random operations use `seed = 42` for reproducibility.*
